# Data Visualization 

The package `ggplot2` is used to visualize data. Before starting you should load the `tidyverse` package, which includes `ggplot2` and other tools, at the very top of your script. 

```{r message = FALSE}
library(tidyverse)
```

## Intro to ggplot2

Read in data

```{r}
df <- read_csv("data/BlueJays.csv")
```

Start with a black canvas 

``` {r}
ggplot()  
```

Just adding the datset to the blank canvas, but nothing plotted yet.

```{r}
ggplot(df)   
```

You need to  use aes() for every ggplot you make, and inside aes() we'll put what our x and y axis will be. 

We say what the x and y are, and it creates scales on each axis, but we didn't tell it what to plot yet.

```{r}
ggplot(df, aes(x=Mass, y=Head) )   
```

Need to add points. 
```{r}
ggplot(df, aes(x=Mass, y=Head) ) + geom_point()
```

What if we want to change the color of the points?
```{r}
ggplot(df, aes(x=Mass, y=Head) ) + geom_point(color='red')
```


What if we want to color the points based on another variable
```{r}
ggplot(df, aes(x=Mass, y=Head, color = KnownSex) ) + geom_point() 
```

Basically, you can just keep adding to the initial plot to make your graph the best visualization to fit your data. 


Another Example!
``` {r}
covid <- read_csv("data/countycovid.csv")

head(covid)
tail(covid)
```

## Histogram 


## Scatter


## Boxplot 


## Adding details 






