# Descriptives

Descriptive statistics describe basic features of the data in simple summaries such as mean, median, and mode. These statistics used to present quantitative descriptions of data in graphing. You can use functions in base R and `tidyverse` to get descriptive statistics. 


```{r echo=FALSE}
library(tidyverse)

atx <- read_csv("data/austin_weather.csv")

head(atx) # first 6 rows
```

Focus in the temperature column for now

```{r}
ggplot(atx, aes(x= temp)) + 
  geom_histogram(color="black", fill="lightseagreen", binwidth = 10)+
  theme_classic()

atx$temp
```

## Basic Descriptives
```{r}
length(atx$temp) # length this tells you the 'n'

sum(atx$temp) # sum

range(atx$temp) # range 

min(atx$temp) # minimum

max(atx$temp) # maximum

var(atx$temp) # variance 

```


## Mean, Median, and Mode 

```{r}
mean(atx$temp) # mean

sum(atx$temp)/length(atx$temp) # mean

median(atx$temp) # median

#estimate mode function 
estimate_mode <- function(x) {
  d <- density(x)
  d$x[which.max(d$y)]
}  

estimate_mode(atx$temp) #estimate mode
```


## Standard Deviation 

``` {r}
sd(atx$temp) #  sample standard deviation

#population standard deviation function 
 pop.sd <- function(s) { 
   sqrt(sum((s - mean(s))^2)/length(s)) 
 } 

pop.sd(atx$temp) # population standard deviation 
```



## Inter-quartile Ranges 

``` {r}
quantile(atx$temp, .25)  # this is the lower quartile

quantile(atx$temp, .75)  # this is the upper quartile

IQR(atx$temp)   # this is the inter-quartile range.

```
these quartiles may be slightly different to what you'd get by hand.

```{r}
ggplot(atx, aes(y=temp)) + 
  geom_boxplot(color='black', fill='lightseagreen')
```

nb we get those numbers on the x-axis because there is no 'group'
we can get rid of them like this:

```{r}
ggplot(atx, aes(y=temp)) + 
  geom_boxplot(color='black', fill='lightseagreen') + 
  scale_x_discrete(breaks = NULL)
```

## Finding 